\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage[hidelinks,colorlinks=true,urlcolor=blue]{hyperref}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, backgrounds, calc}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

% TikZ styles for the architecture diagram
\tikzset{
    frozen/.style   = {rectangle, rounded corners, draw=blue!60, fill=blue!8,
                       minimum width=2.0cm, minimum height=0.85cm, align=center,
                       font=\footnotesize},
    trainable/.style= {rectangle, rounded corners, draw=red!70, fill=red!8,
                       minimum width=2.2cm, minimum height=0.85cm, align=center,
                       font=\footnotesize},
    gate/.style     = {diamond, aspect=2, draw=orange!80, fill=orange!10,
                       minimum width=2.2cm, minimum height=0.6cm, align=center,
                       font=\footnotesize, inner sep=1pt},
    inputbox/.style = {rectangle, draw=black!50, fill=gray!8,
                       minimum width=1.7cm, minimum height=0.7cm, align=center,
                       font=\footnotesize},
    output/.style   = {rectangle, rounded corners, draw=green!60!black, fill=green!10,
                       minimum width=2.0cm, minimum height=0.7cm, align=center,
                       font=\footnotesize},
    ttt/.style      = {rectangle, rounded corners, draw=purple!70, fill=purple!8,
                       minimum width=3.0cm, minimum height=1.6cm, align=center,
                       font=\footnotesize},
    flow/.style     = {-Stealth, thick, draw=black!70},
    skipflow/.style = {-Stealth, thick, draw=green!50!black, dashed},
    adaptflow/.style= {-Stealth, thick, draw=purple!70, dashed},
    note/.style     = {font=\scriptsize\itshape, text=black!60}
}

\begin{document}

\title{AdaTTT: Adaptive Test-Time Training for\\
Efficient Visual Question Answering}

\author{
\IEEEauthorblockN{Yugesh Reddy}
\IEEEauthorblockA{\textit{Sappidi}\\
                UIN: }
\and
\IEEEauthorblockN{Aishwarya Reddy}
\IEEEauthorblockA{\textit{Chinthalapudi}\\
        UIN: 665635004}
\and
\IEEEauthorblockN{Aryan}
\IEEEauthorblockA{\textit{Shetty}\\
        UIN: }
}

\maketitle

\begin{abstract}
Visual Question Answering (VQA) models receive an image and a natural language
question and predict an answer. These models often degrade when the test
distribution differs from the training distribution, such as on blurry photos,
low-quality images, unusual objects, or OCR-heavy memes. Test-Time Training
(TTT) addresses this by performing a few self-supervised steps on each test
sample before prediction, but applying TTT to every input is expensive and
frequently harms easy examples. We present \textbf{AdaTTT}, a framework that
learns \emph{when} adaptation is needed. AdaTTT combines a frozen ViT-B/16
vision encoder, a frozen BERT-base text encoder, a trainable bidirectional
cross-modal fusion module, a prediction head, and a lightweight confidence
gate that routes each sample to either a fast skip path or a TTT-enabled
adapt path. Across the saved 214{,}354-sample VQA-v2 validation set, naïve
always-on TTT reduces accuracy from \textbf{49.56\%} (base) to \textbf{48.49\%}
at $K{=}1$ while inflating compute by 40\%; further steps degrade accuracy
further. AdaTTT, by contrast, recovers \textbf{49.56\%} accuracy at threshold
$\tau{=}0.9$ while routing only 0.96\% of samples through TTT, yielding a
flat compute profile. On Memotion2, the same architecture reaches
\textbf{71.65\%} validation accuracy with a balanced 50/50 skip/adapt split,
demonstrating cross-task transfer. The contribution is a systems result:
selective adaptation can contain the harm of indiscriminate TTT while
preserving accuracy and compute near the base model.\\
\textbf{Code and artifacts:}
\url{https://drive.google.com/drive/folders/1Bs9AXg5yVbtT9In8t2FfW4bA9n3mf0J_?usp=share_link}
\end{abstract}

\begin{IEEEkeywords}
Test-time training, visual question answering, multimodal learning,
adaptive inference, confidence gating, distribution shift.
\end{IEEEkeywords}

\section{Introduction}

Visual Question Answering (VQA) is a multimodal task where a model receives
an image and a question and produces an answer drawn from a fixed
vocabulary. For example, given an image of a red bus and the question
``What color is the bus?'', the model should answer ``red.'' Modern VQA
systems require both visual understanding and language understanding, and
typically rely on large pretrained encoders fused through a multimodal
backbone.

Deep models perform well when the test distribution matches training data,
but real-world inputs often violate this assumption. Test photos may have
poor lighting, motion blur, novel objects, or text-heavy content such as
memes. This mismatch, broadly termed \emph{distribution shift}, degrades
the predictions of frozen models because the features they extract no
longer align with the decision boundaries they learned. \emph{Test-Time
Training} (TTT) is one remedy~\cite{b6}: before each prediction, the model
takes a few self-supervised gradient steps using the test input itself,
adjusting selected internal parameters so that they better fit \emph{this
particular} input before the final answer is produced. The hope is that
even a handful of inner-loop updates can correct the visual encoder or
the multimodal fusion just enough to recover the right answer on a
hard sample.

The downside is that TTT is expensive and not universally helpful. Each
adapted sample pays the cost of $K$ extra forward-backward passes through
the trainable modules, plus the bookkeeping for parameter save and
restore. Worse, TTT can actively disturb \emph{correct} predictions on
easy samples: the inner-loop optimizer has no notion of whether the
current input is hard or easy, so on samples the base model already
handles well, the SSL gradients can drift the representation off the
manifold the prediction head was trained for. Applying TTT to every
test sample therefore wastes compute on the easy cases and inflicts a
small but measurable accuracy cost on them. Our saved VQA-v2 measurements
show this cost concretely: always-on TTT at $K{=}1$ drops accuracy from
49.56\% to 48.49\% and grows compute by 40\%; at $K{=}5$ accuracy falls
to 46.48\% and compute roughly triples.

\textbf{The central question of this project is therefore:} \emph{can we
make test-time training selective, cheap enough, and safe enough that it
only runs when it is likely to help?} Equivalently: can we treat TTT
not as a fixed inference policy applied to every sample, but as a
\emph{conditional resource} allocated only to inputs whose representation
actually needs sharpening?

AdaTTT answers this with a confidence-gated architecture. A small learned
gate scores each fused multimodal representation; high-confidence samples
take the fast \emph{skip} path and are answered directly, while
low-confidence samples take the slow \emph{adapt} path through TTT. The
gate is trained with labels derived from \emph{actually} comparing base
and TTT predictions, not just from raw confidence. The result is a system
that produces a Pareto frontier of accuracy versus compute rather than a
single static operating point.

\subsection{Problem Statement}

Given an input image $I$ and question $Q$, the model predicts an answer
$\hat{y} \in \{1,\dots,N\}$ from a fixed vocabulary. For VQA-v2, $N{=}3{,}129$
covers the most frequent answers. For Memotion2, $N{=}3$ covers the
sentiment classes (positive, negative, neutral). Standard always-on TTT
applies a self-supervised inner loop to every test input, but this is
inefficient on easy samples and may disturb the model's learned image-text
alignment. AdaTTT replaces this uniform policy with a learned routing
function $g(\cdot;\theta_g)$ that sends each sample to one of two paths
based on its predicted likelihood of benefiting from adaptation.

\subsection{Contributions}

\begin{itemize}
    \item A confidence-gated multimodal pipeline with bidirectional cross-modal
    fusion, frozen pretrained encoders, and per-sample save-and-restore TTT
    semantics.
    \item A two-phase gate training procedure that refines the gate against
    \emph{outcome-based} labels comparing base and TTT predictions, rather
    than treating raw model confidence as a routing signal.
    \item A complete experimental program including TTT step sweeps,
    threshold sweeps, stabilization ablations, latency profiling, and
    cross-task transfer to Memotion2.
    \item A clear empirical finding: indiscriminate TTT is harmful in this
    saved VQA setting, and selective routing recovers baseline accuracy
    while keeping compute near the base path.
\end{itemize}

\section{Related Work}

\textbf{Test-Time Training.} Sun \emph{et~al.}~\cite{b6} introduced TTT as
a way to adapt a model to each test sample using a self-supervised
auxiliary task. The original formulation jointly trains a main task head
$\theta_d$ and a self-supervised head $\theta_s$ on a shared feature
extractor $\theta_f$ during supervised pretraining, and then at test time
performs a small number of gradient steps on $\{\theta_f, \theta_s\}$
using only the SSL loss before producing the final main-task prediction.
The premise is that if the SSL task and the main task share a feature
extractor, improving SSL performance on a single test input should also
sharpen the features that the main task consumes. Subsequent work
explored entropy minimization, batch-norm statistics adaptation, and
prompt tuning as alternative inner-loop signals. Our work treats TTT as a
tool to be invoked \emph{selectively} rather than uniformly, and pairs it
with explicit save-and-restore semantics to prevent cross-sample
contamination---two design choices that the original formulation does
not require but that become essential once TTT is wrapped inside a
production-style adaptive router.

\textbf{Visual Question Answering.} VQA-v2 is the canonical benchmark
for image-question reasoning~\cite{b1}. VizWiz adds natural distribution
shift through photographs taken by blind users~\cite{b2}. Memotion2
extends multimodal evaluation to memes and sentiment~\cite{b3}.

\textbf{Pretrained encoders.} ViT~\cite{b4} and BERT~\cite{b5} are widely
used as frozen feature extractors. We adopt this freezing strategy not
just for cost reasons but because it makes feature caching practical and
enables modular cross-task transfer.

\section{Datasets}

\textbf{VQA-v2}~\cite{b1} is the primary dataset. It contains image-question
-answer triples grounded in MS-COCO images, with three major question
types (\texttt{yes/no}, \texttt{number}, \texttt{other}) and 3{,}129 answer
classes covering the most frequent responses. The pipeline operates over
443{,}757 training questions and 214{,}354 validation questions. The
dataset has strong language priors, which makes it a good probe for when
visual adaptation \emph{actually} helps.

\textbf{VizWiz}~\cite{b2} contains questions over photographs taken by
blind users. Images are noisier, off-axis, and frequently include
``unanswerable'' as a class, providing natural distribution shift. The
loader is implemented in this repository, but no large saved VizWiz run
is included in the current snapshot.

\textbf{Memotion2}~\cite{b3} is used for cross-task transfer. It contains
meme images paired with OCR-extracted text, with sentiment labels mapped
into three classes. The pipeline reuses ViT and BERT, the fusion module,
and the gate; only the prediction head's output dimension changes. The
saved run uses 11{,}187 training and 2{,}797 validation samples; the
loader collapses Memotion2's finer-grained sentiment labels
(\texttt{very\_positive}, \texttt{very\_negative}) into the neutral
bucket, which is a known caveat when interpreting absolute accuracy.

\section{Background: Test-Time Training in Depth}
\label{sec:ttt-bg}

This section gives a self-contained explanation of what test-time training
actually does, why it can help, and---equally importantly---why it can
hurt. The arguments here motivate the central design decision of AdaTTT:
TTT must be \emph{conditional}, not unconditional.

\subsection{Why a Frozen Model Fails Under Distribution Shift}

A trained classifier $f_{\theta} : \mathcal{X} \to \mathcal{Y}$ partitions
its input space into decision regions whose boundaries are calibrated
against the empirical training distribution $p_{\mathrm{train}}(x,y)$.
At test time we draw inputs from some $p_{\mathrm{test}}(x)$; whenever
$p_{\mathrm{test}} \ne p_{\mathrm{train}}$ the input embeddings can land
in regions where the boundary was never well-calibrated. In a multimodal
VQA system three sources of shift compound:

\begin{itemize}
\item \textbf{Visual shift.} Lighting, camera pose, blur, noise, and
unusual objects move ViT features off their training manifold.
\item \textbf{Lexical/syntactic shift.} New question phrasings or OCR
text from memes can land outside BERT's training distribution.
\item \textbf{Cross-modal binding shift.} Even when individual encoders
generalize, the \emph{joint} distribution of $(v, t)$ pairs at test time
can differ from training, breaking the alignment learned by the fusion
module.
\end{itemize}

Static fine-tuning cannot fix these on a per-sample basis because every
shift is unique to the specific test input. This is the regime where
test-time adaptation, in principle, has something to offer.

\subsection{What TTT Actually Does, Mechanically}

For a single test input $x_t$, classical TTT performs the inner loop:
\begin{equation}
\theta^{(i+1)} = \theta^{(i)} - \eta\, \nabla_{\theta}\,
\mathcal{L}_{\mathrm{ssl}}\bigl(x_t;\, \theta^{(i)}\bigr),\quad i=0,\dots,K-1
\end{equation}
where $\theta$ is a chosen subset of parameters and $\mathcal{L}_{\mathrm{ssl}}$
is a self-supervised loss whose label is derived from $x_t$ itself
(masked-patch reconstruction, rotation prediction, contrastive consistency,
entropy, etc.). The main-task prediction is then taken with the adapted
parameters: $\hat{y}_t = \arg\max f(x_t; \theta^{(K)})$.

Two implicit assumptions make this work \emph{when} it works:

\begin{enumerate}
\item \textbf{Feature-sharing assumption.} The features whose update
direction is set by $\nabla \mathcal{L}_{\mathrm{ssl}}$ are also features
the main task depends on. If the SSL objective sharpens visual encoding,
and the main classifier consumes those visual features, the main task
benefits.
\item \textbf{Local-update assumption.} A small number of inner-loop
steps on a single sample do not move $\theta$ far enough to catastrophically
forget the supervised prior. The adapted model is still a tiny perturbation
of the base model.
\end{enumerate}

Both assumptions are fragile. The first fails when the SSL objective
learns a spurious shortcut (e.g.\ rotation can sometimes be predicted
from low-frequency colour statistics that the main task does not use).
The second fails when the SSL gradient's magnitude or direction conflicts
with what the main-task loss would have wanted---in which case TTT
\emph{actively} steers the model toward worse main-task predictions.

\subsection{Why Per-Sample TTT Needs Save-and-Restore}

A subtle but critical point: if we leave the adapted parameters in place
and move on to the next test sample, the inner-loop updates
\emph{accumulate}. Because each update is computed on a single noisy
SSL signal, the accumulated drift behaves like a random walk in
parameter space whose step size is much larger than the gradient noise
seen during ordinary minibatch training. Empirically this drift is
catastrophic: in our pilot runs, leaving TTT updates persistent collapsed
validation accuracy after roughly 100 samples. Restoring $\theta_f,
\theta_d$ after each sample is therefore not a stylistic choice but a
correctness invariant. Algorithm~\ref{alg:ttt} encodes this; the test
suite verifies it on both success and exception paths.

\subsection{Why Naïve Always-On TTT Fails on VQA}
\label{sec:ttt-fails}

Three independent failure modes combine to make uniform TTT a poor
inference policy on VQA-v2:

\textbf{(a) Most VQA samples do not need adaptation.} VQA-v2 has strong
language priors~\cite{b1}: a substantial fraction of yes/no and number
questions can be answered correctly using lexical patterns alone. For
these inputs the base model is already correct, and any per-sample
gradient update is at best wasted compute and at worst harmful
perturbation.

\textbf{(b) The SSL signal is misaligned with the answer signal.} Both
masked-patch reconstruction and rotation are visual SSL objectives. They
nudge $\theta_f$ to produce features that better support image
self-supervision, which is not the same as features that better support
the cross-modal answer prediction. When the two objectives disagree,
the inner loop drifts the fusion away from the alignment learned during
supervised training. This is exactly the failure pattern visible in
Table~\ref{tab:base-vs-ttt}: more $K$ amplifies the drift, accuracy
falls monotonically.

\textbf{(c) The compute cost is paid every sample, regardless of value.}
Standard TTT charges $\sim$3.2 GFLOPs per inner step on top of the
$\sim$40.9 GFLOPs base path. At $K{=}1$ that is a 40\% surcharge; at
$K{=}5$ it is a 3$\times$ surcharge. If the marginal accuracy gain on
the few samples where TTT helps is smaller than the marginal accuracy
loss on the many samples where TTT hurts, the average is negative
\emph{and} the deployment is paying the worst possible cost-to-benefit
ratio.

The combined picture is that always-on TTT pays a uniform tax on every
input but only delivers value to a small subset of them. The remedy is
not a better SSL loss or a larger $K$; it is to stop paying the tax on
samples that do not need it.

\subsection{Why Adaptive TTT Is the Right Solution}
\label{sec:why-adaptive}

The argument has four pieces: workload skew, asymmetric cost, decision-
theoretic optimality, and empirical evidence.

\textbf{Workload skew.} On VQA-v2, the saved 50{,}000-sample gate-label
study shows that only \textbf{1.01\%} of training samples are
\emph{TTT-helps} cases (base wrong, TTT right) versus \textbf{98.99\%}
\emph{skip} cases. If only 1\% of samples genuinely benefit from
adaptation, any inference policy that treats all samples uniformly is
spending 99\% of its TTT budget on inputs that do not need it. A
selective policy can recover essentially the same gains for 1/99 of the
cost---if it can identify the right 1\%.

\textbf{Asymmetric cost.} Skipping is cheap (one base forward pass) and
has known accuracy. Adapting is expensive (one base pass + $K$ inner-loop
steps + restore) and has uncertain accuracy that may be better, the
same, or worse than skipping. A risk-averse inference policy should
adapt only when the expected gain exceeds the expected cost. Concretely,
let $p_h$ be the probability that adapting helps a particular sample,
$p_n$ that it is neutral, and $p_d$ that it damages the prediction; let
$\Delta_+$ and $\Delta_-$ be the magnitudes of the help and harm. The
expected accuracy change of adapting is $p_h\Delta_+ - p_d\Delta_-$. On
VQA-v2 the saved data say $p_h \approx 0.01$ and $p_d$ is non-trivial,
so the unconditional expected change is negative---which is exactly
what Table~\ref{tab:base-vs-ttt} measures.

\textbf{Decision-theoretic optimality.} The optimal policy minimizes
expected loss conditional on each input, not unconditionally. If we had
oracle access to a function $\mathrm{helps}(x) \in \{0,1\}$ telling us
whether TTT improves $x$, the right policy would be $\pi^*(x) =
\mathrm{adapt}$ iff $\mathrm{helps}(x) = 1$. The confidence gate is a
\emph{learned approximation} to this oracle, trained directly against
the outcome labels described in Section~\ref{sec:training}. This is why
the gate is trained on labels derived from comparing actual base versus
TTT predictions, not on raw model confidence: confidence answers ``how
sure is the base model?'' but the policy needs the answer to ``does
adaptation change the base prediction in the right direction?''.

\textbf{Empirical evidence.} The saved gate sweep
(Table~\ref{tab:gate-sweep}) shows that AdaTTT at $\tau{=}0.9$ recovers
\textbf{49.56\%} accuracy at \textbf{46.49} GFLOPs/sample, versus base
49.56\% at 46.31 GFLOPs and always-on TTT at 48.49\% / 64.91 GFLOPs.
That is essentially \emph{full base accuracy at essentially base
compute}, while still leaving a small TTT escape valve open for the
0.96\% of samples the gate flagged. Equally importantly, the same
machinery on Memotion2 routes 49.9\% of samples through TTT---showing
that the gate is not a hard-coded skip and that the system genuinely
allocates adaptation where the data signals it is useful.

In short: TTT is a tool with a steep cost curve and a narrow help window.
The right way to deploy it is conditionally, against a learned routing
function trained on actual outcomes. That is what AdaTTT does.

\subsection{Comparison With Alternative Inference Strategies}
\label{sec:alternatives}

It is worth contrasting adaptive TTT with the obvious alternatives:

\begin{itemize}
\item \textbf{Static fine-tuning} on the test domain requires labels and
adapts a single global $\theta$ rather than per-sample. It cannot
respond to individual hard inputs at all.
\item \textbf{Always-on TTT} pays the full cost on every sample and, as
shown above, suffers a net loss on workloads where most samples are
easy.
\item \textbf{Confidence-thresholded refinement} (skip if base softmax
$> \tau_{\mathrm{conf}}$) is a tempting heuristic but uses the
\emph{wrong signal}. Base softmax measures internal certainty about an
answer, not the probability that adaptation would change the answer for
the better. Outcomes-based gate labels capture that probability directly.
\item \textbf{Test-time entropy minimization} adapts using prediction
entropy as the signal and is therefore subject to the same uniform-cost
problem and to confirmation-bias drift on confident-but-wrong samples.
\item \textbf{Adaptive TTT} (this work) decouples \emph{when} from
\emph{how}. The gate decides when, and standard SSL TTT handles how.
The two sub-problems can then be improved independently.
\end{itemize}

This decoupling is what gives AdaTTT its Pareto behavior: tuning $\tau$
sweeps a smooth curve from base-equivalent compute to always-on TTT,
and the user can pick any operating point along it.

\section{Method}

\subsection{System Overview}

Figure~\ref{fig:arch} shows the full AdaTTT pipeline. An image $I$ and
text $Q$ are encoded by frozen ViT-B/16 and BERT-base into token sequences,
which a 2-layer bidirectional cross-modal fusion module combines into a
pooled representation $z \in \mathbb{R}^{768}$. A small confidence gate
$g(\cdot)$ scores this representation. If $g(z) > \tau$, the sample takes
the fast skip path: the prediction head produces $\hat{y}$ directly. If
$g(z) \le \tau$, the sample enters the TTT loop, where the fusion module
$\theta_f$ and prediction head $\theta_d$ are adapted for $K$ gradient
steps using a self-supervised loss; the model predicts $\hat{y}$ with the
adapted parameters and then restores $\theta_f, \theta_d$ to their pre-
adaptation state. The encoders and the gate are never updated at test
time. A graceful-degradation fallback layer wraps the whole pipeline,
catching TTT timeouts, OOM events, and image-loading failures.

\begin{figure*}[!t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[node distance=0.9cm and 1.1cm]

% Inputs
\node[inputbox] (img) {Image $I$\\(224$\times$224)};
\node[inputbox, below=0.6cm of img] (q) {Question $Q$\\(BERT tokens)};

% Frozen encoders
\node[frozen, right=1.2cm of img] (vit) {Frozen ViT-B/16\\$v\!\in\!\mathbb{R}^{197\times 768}$};
\node[frozen, right=1.2cm of q]   (bert){Frozen BERT-base\\$t\!\in\!\mathbb{R}^{L\times 768}$};

% Fusion
\node[trainable, right=1.4cm of vit, yshift=-0.9cm,
      minimum width=2.6cm, minimum height=1.6cm] (fuse)
      {Cross-Modal Fusion $\theta_f$\\
       2-layer bidirectional\\
       cross-attention\\
       $z\!\in\!\mathbb{R}^{768}$};

% Gate
\node[gate, right=1.4cm of fuse] (gate) {Confidence Gate $\theta_g$\\$g(z) \gtrless \tau$?};

% Skip path -> Pred head
\node[trainable, above right=0.4cm and 1.6cm of gate] (head1)
      {Prediction Head $\theta_d$\\$\mathbb{R}^{768}\!\to\!\mathbb{R}^{N}$};

% Adapt path -> TTT block -> Pred head
\node[ttt, below right=0.4cm and 1.0cm of gate] (ttt) {%
   TTT Inner Loop ($K$ steps)\\
   1.\ Save anchor $\theta_f^*,\theta_d^*$\\
   2.\ Self-sup.\ loss (masked\\
   \ \ \ patch $\mid$ rotation)\\
   3.\ Adam step on $\theta_f,\theta_d$\\
   4.\ Predict $\hat y$\\
   5.\ \textbf{Restore} $\theta_f^*,\theta_d^*$};

\node[trainable, right=1.0cm of ttt] (head2)
      {Prediction Head $\theta_d$\\(adapted)};

% Outputs
\node[output, right=1.2cm of head1] (yhat1) {$\hat{y}$ (skip)};
\node[output, right=1.2cm of head2] (yhat2) {$\hat{y}$ (adapt)};

% Flows: inputs -> encoders
\draw[flow] (img) -- (vit);
\draw[flow] (q) -- (bert);

% Encoders -> fusion
\draw[flow] (vit.east)  -- ++(0.4,0) |- (fuse.west);
\draw[flow] (bert.east) -- ++(0.4,0) |- (fuse.west);

% Fusion -> gate
\draw[flow] (fuse) -- (gate);

% Gate -> skip path
\draw[skipflow] (gate.north) |- node[note, near start, above]
      {$g(z)\!>\!\tau$ (skip)} (head1.west);

% Gate -> adapt path
\draw[adaptflow] (gate.south) |- node[note, near start, below]
      {$g(z)\!\le\!\tau$ (adapt)} (ttt.west);

% Adapt path through TTT
\draw[flow] (ttt) -- (head2);

% Heads -> outputs
\draw[flow] (head1) -- (yhat1);
\draw[flow] (head2) -- (yhat2);

% Background grouping: TTT updates fusion+head
\begin{scope}[on background layer]
   \node[draw=purple!30, dashed, rounded corners,
         fit=(fuse) (ttt) (head2),
         inner sep=10pt, label={[note,purple!70]above:Updated during TTT}]
        {};
\end{scope}

% Legend
\node[note, below=1.6cm of ttt, anchor=west, xshift=-3.5cm]
   (lg) {%
   \begin{tabular}{@{}l@{\ }l@{\quad}l@{\ }l@{\quad}l@{\ }l@{\quad}l@{\ }l@{}}
   \tikz\node[frozen,minimum width=0.55cm,minimum height=0.30cm]{}; & Frozen
   & \tikz\node[trainable,minimum width=0.55cm,minimum height=0.30cm]{}; & Trainable
   & \tikz\node[gate,minimum width=0.55cm,minimum height=0.30cm]{}; & Routing
   & \tikz\node[ttt,minimum width=0.55cm,minimum height=0.30cm]{}; & TTT loop
   \end{tabular}};

\end{tikzpicture}%
}
\caption{AdaTTT architecture flow. Frozen ViT-B/16 and BERT-base encode
image and text into token sequences. A trainable bidirectional cross-modal
fusion module produces a pooled vector $z$. The confidence gate routes
$z$ either to the prediction head directly (skip path, dashed green) or
through a per-sample TTT inner loop that adapts $\theta_f, \theta_d$ and
restores them after prediction (adapt path, dashed purple). The gate
itself is never updated at test time.}
\label{fig:arch}
\end{figure*}

\subsection{Frozen Encoders}

The vision encoder is \texttt{google/vit-base-patch16-224} and the text
encoder is \texttt{bert-base-uncased}; both have \texttt{requires\_grad
=False} and run inside \texttt{torch.no\_grad()}. Freezing has three
benefits. First, it bounds the trainable problem to a small set of
parameters (Section~\ref{sec:params}), which keeps both standard training
and per-sample TTT tractable. Second, frozen forward passes are
deterministic for a given input, which enables the cached-features
optimization (Section~\ref{sec:cache}). Third, modularity at the encoder
boundary makes the same backbone usable for VQA, VizWiz, and Memotion2.

\subsection{Cross-Modal Fusion}

The fusion module $\theta_f$ is a two-layer bidirectional cross-modal
attention block. Visual tokens query text (Q$=$visual, K$=$V$=$text), and
text tokens then query visual tokens, with LayerNorm and feed-forward
sublayers between rounds. The output is a pooled 768-d vector $z$. This
representation is consumed by both the gate and the prediction head, and
is the locus of TTT updates: any improvement from per-sample adaptation
manifests here.

\subsection{Confidence Gate}

The gate $\theta_g$ is a small MLP
$\mathrm{Linear}(768{\to}256)\!\to\!\mathrm{ReLU}\!\to\!\mathrm{Dropout}(0.1)
\!\to\!\mathrm{Linear}(256{\to}1)\!\to\!\sigma$. Its routing rule is
\[
\text{path}(z) = \begin{cases}
\text{skip}  & \text{if } g(z) > \tau, \\
\text{adapt} & \text{if } g(z) \le \tau.
\end{cases}
\]
Critically, the gate is \emph{frozen at test time}. Its learning
happens in two distinct phases (Section~\ref{sec:training}): a weak
proxy phase during base training, and an outcome-driven refinement
phase against gate labels.

\subsection{Prediction Head}

The prediction head $\theta_d$ is
$\mathrm{Linear}(768{\to}1024)\!\to\!\mathrm{ReLU}\!\to\!\mathrm{Dropout}
(0.2)\!\to\!\mathrm{LayerNorm}\!\to\!\mathrm{Linear}(1024{\to}N)$.
For VQA, $N{=}3{,}129$. For Memotion2, $N{=}3$.

\subsection{Test-Time Training Loop}
\label{sec:ttt-loop}

For every sample routed to \emph{adapt}, AdaTTT executes the inner
loop in Algorithm~\ref{alg:ttt}. This subsection unpacks each step of
that algorithm, since the precise mechanics matter for both correctness
and reproducibility.

\begin{algorithm}[!t]
\caption{Per-sample TTT with save-and-restore}
\label{alg:ttt}
\begin{algorithmic}[1]
\REQUIRE Sample $(I,Q)$, params $\theta_f, \theta_d$, steps $K$, lr $\eta$
\STATE $\theta_f^* \leftarrow \mathrm{clone}(\theta_f);\quad \theta_d^* \leftarrow \mathrm{clone}(\theta_d)$
\STATE $\mathrm{opt}\leftarrow \mathrm{Adam}(\{\theta_f,\theta_d\}, \mathrm{lr}{=}\eta)$
\FOR{$i = 1$ \TO $K$}
   \STATE $\mathcal{L}_{\mathrm{ssl}} \!\leftarrow\! \mathcal{L}_{\mathrm{mask}}(I,Q;\theta_f,\theta_d)$ or $\mathcal{L}_{\mathrm{rot}}(I;\theta_f,\theta_d)$
   \STATE $\mathrm{opt.zero\_grad}();\ \mathcal{L}_{\mathrm{ssl}}.\mathrm{backward}()$
   \STATE $\mathrm{clip\_grad}(1.0);\ \mathrm{opt.step}()$
\ENDFOR
\STATE $\hat{y} \leftarrow \mathrm{argmax}\;f(I,Q;\theta_f,\theta_d)$
\STATE $\theta_f \leftarrow \theta_f^*;\quad \theta_d \leftarrow \theta_d^*$ \COMMENT{\textbf{restore}}
\RETURN $\hat{y}$
\end{algorithmic}
\end{algorithm}

\noindent\textbf{Step-by-step.}

\textit{Lines 1--2 (anchor + fresh optimizer).} The current values of
$\theta_f$ and $\theta_d$ are deep-copied into anchor tensors
$\theta_f^*, \theta_d^*$. Adam state is created from scratch every
adaptation episode; the running first- and second-moment estimates from
the supervised training optimizer are intentionally not reused, because
they encode population-level statistics that have no meaning for a
single-sample inner loop. Reusing the training optimizer would also
leak per-sample momentum across samples, which would silently break the
per-sample independence guarantee.

\textit{Lines 3--6 (inner loop).} The SSL objective is computed on the
\emph{current} test input only---no labels, no ground-truth answers, no
information that would not be available at deployment. The backward pass
flows through fusion and prediction head only; the encoders are wrapped
in \texttt{torch.no\_grad()} so their parameters never appear in the
optimizer. Gradients are clipped to global norm 1.0 before each step;
this is more important here than in supervised training because
single-sample SSL gradients have higher variance than minibatch averages
and a single noisy step can otherwise overshoot drastically.

\textit{Line 7 (adapted prediction).} The forward pass that produces
$\hat{y}$ uses $\theta_f^{(K)}, \theta_d^{(K)}$ rather than any
intermediate. Predicting at the final inner-loop iterate, not at an
average, keeps the semantics of ``$K$ steps of adaptation'' precise.

\textit{Line 8 (restore).} \emph{This is the most important line of the
algorithm.} A direct \texttt{param.data.copy\_(saved\_state)} replaces
both $\theta_f$ and $\theta_d$ with their pre-adaptation snapshots. The
optimizer object is discarded with the function scope. After this line
the model is bit-identical to its pre-adaptation state, so the next
sample's adaptation episode begins from a clean slate.

\textbf{Why this structure matters.} The inner loop is small enough
($K \le 5$) that it cannot accumulate enough signal to override the
supervised pretraining. That is exactly the point: TTT is supposed to
nudge the representation toward a better fit for one input, not retrain
the model. The save-and-restore protocol turns a stateful sequence of
gradient updates into a stateless function of a single input, which is
the only semantics under which per-sample adaptation can be meaningfully
evaluated. Without restoration, every reported accuracy number would
silently depend on test-set ordering, which is not a property a
research result should have.

\textbf{Which parameters get adapted.} Three principled choices govern
the trainable subset $\{\theta_f, \theta_d\}$:

\begin{itemize}
\item \emph{Encoders are excluded.} Adapting ViT or BERT during a
single-sample inner loop would risk catastrophic forgetting of their
pretraining, and would also negate the cached-features optimization.
\item \emph{The gate is excluded.} The gate's whole purpose is to make
a stable \emph{a priori} decision about routing; updating it during the
adapt path would create a circular dependency where the routing changes
mid-decision.
\item \emph{Fusion and head are included.} These are the two modules
where multimodal alignment lives, and they are also the smallest
trainable surface that can plausibly absorb a per-sample correction.
The component ablation script confirms that excluding either degrades
behavior.
\end{itemize}

\subsection{Self-Supervised Objectives}
\label{sec:ssl}

Two TTT objectives are implemented; both are visual SSL tasks because
the visual encoder is where most distribution shift in VQA actually
manifests.

\textbf{Masked patch reconstruction (mainline).} The image is encoded
into 197 ViT tokens (one CLS token plus 196 patch tokens for a
$14{\times}14$ grid). A fraction $\rho{=}0.25$ of the patch tokens
(indices $1$--$196$) is randomly selected and masked. A small linear
head $h_{\mathrm{mask}}: \mathbb{R}^{768} \to \mathbb{R}^{768}$ predicts
the \emph{mean of the masked tokens} from the pooled fused representation
$z$, and the loss is the mean-squared error against the ground-truth
mean of the unmasked features:
\begin{equation}
\mathcal{L}_{\mathrm{mask}} = \bigl\| h_{\mathrm{mask}}(z) -
\mathrm{mean}\bigl\{ v_j : j \in \mathcal{M} \bigr\} \bigr\|_2^2.
\end{equation}
The intuition is that to predict masked-region statistics from the
pooled multimodal vector, the fusion module must propagate enough
visual information through to $z$. Adapting under this loss therefore
sharpens visual encoding \emph{conditioned on} the question, not in
isolation. This is why masked-patch is the mainline objective: it
implicitly couples to the multimodal task even though its supervision
signal is purely visual.

\textbf{Rotation prediction (alternative).} The input image is rotated
by an angle drawn from $\{0^\circ, 90^\circ, 180^\circ, 270^\circ\}$,
re-encoded through the frozen ViT, and a 4-way classifier predicts the
rotation class from $z$ with cross-entropy loss. Rotation is a simpler
SSL task and does not depend on the fusion's ability to surface visual
detail; that is also its weakness. Empirically, rotation TTT
underperforms masked-patch TTT on VQA-v2 by 1.30~pt at $K{=}1$
(Table~\ref{tab:base-vs-ttt}), which is consistent with the hypothesis
that rotation's signal is too coarse to drive useful per-sample updates
in a multimodal setting.

\textbf{Why two objectives.} Implementing both lets us distinguish
``TTT failed'' from ``this particular SSL signal failed.'' The fact
that masked-patch beats rotation but neither beats base in the
always-on regime points cleanly at the workload-skew problem
(Section~\ref{sec:why-adaptive}) rather than an objective-design
problem.

\subsection{Stabilization Techniques}

Per-sample adaptation can drift in harmful directions. Two stabilizers
are implemented and ablated:

\begin{itemize}
    \item \textbf{Consistency regularization.} Two augmented views
    (\texttt{RandomResizedCrop} + \texttt{ColorJitter}) are fused
    independently; a symmetric KL divergence between their prediction
    distributions is added to the TTT loss with weight $\lambda_{\mathrm{
    cons}}{=}0.1$.
    \item \textbf{Mixup anchoring.} The adapted representation is
    interpolated toward its pre-adaptation anchor:
    $z_{\mathrm{mix}} = \alpha\,z_{\mathrm{cur}} + (1-\alpha)\,z^*$ with
    $\alpha \in [0.7, 1.0]$ and $z^*$ detached. This bounds drift while
    permitting useful updates.
\end{itemize}

\subsection{Adaptive Routing and FLOPs Accounting}

At inference, each batch is split by the gate into a skip subset and an
adapt subset, processed independently, and recombined in original order.
Per-sample FLOPs are tracked: a skip sample costs roughly
$F_{\mathrm{skip}} \approx 40.9$ GFLOPs (encode + fuse + predict), and
an adapt sample costs $F_{\mathrm{adapt}} \approx 40.9 + K\cdot 3.2$
GFLOPs. Average compute is the skip-rate-weighted mixture of the two,
which is what makes the Pareto frontier non-trivial.

\subsection{Cached Frozen Features}\label{sec:cache}

Because ViT and BERT are frozen, their outputs are deterministic per
sample. The pipeline therefore precomputes encoder outputs once and
persists them across TTT sweeps, gate sweeps, ablations, and adaptive
inference. The saved cache is sharded ($\sim$16 GB shards for the
validation split), which is necessary because the full validation cache
exceeds reasonable single-file memory budgets. This optimization makes
repeated experiments tractable.

\subsection{Graceful Degradation Fallback}

A four-level fallback wraps the adaptive router:
(0) full AdaTTT, (1) base-only on TTT timeout/error,
(2) reduced-resolution retry on OOM, (3) hard-error result. Levels are
encoded as an \texttt{IntEnum} so latency budgets and routing decisions
remain analyzable end-to-end.

\section{Training Pipeline}\label{sec:training}

AdaTTT is trained in three sequential phases.

\subsection{Phase 1: Base Supervised Training}

The fusion module, prediction head, and gate are trained jointly under a
dual loss
\[
\mathcal{L}_{\mathrm{base}} = \mathrm{CE}(\mathrm{logits}, y)
+ 0.1\cdot \mathrm{BCE}\bigl(g(z),\, \mathbb{1}[\hat{y}{=}y]\bigr).
\]
The BCE term gives the gate a weak proxy signal: ``predict whether the
base model gets this sample right.'' Optimizer: AdamW, lr $1\mathrm{e}{-4}$,
weight decay $0.01$, cosine schedule with 10\% warmup, batch size 64,
15 epochs.

\subsection{Phase 2: Oracle Gate-Label Generation}

After base training, both the base model and the TTT model are run on a
50{,}000-sample subset of the training split. Outcomes are mapped to gate
labels:

\begin{center}\footnotesize
\begin{tabular}{lll}
\toprule
Base correct & TTT correct & Gate label \\
\midrule
yes & yes & $1.0$ (skip) \\
yes & no  & $1.0$ (skip; TTT hurts) \\
no  & yes & $0.0$ (adapt; TTT helps) \\
no  & no  & $0.5$ (uncertain, masked) \\
\bottomrule
\end{tabular}
\end{center}

\noindent Of the 48{,}099 retained labels, 47{,}612 (98.99\%) say
\emph{skip} and only 487 (1.01\%) say \emph{adapt}. This skew is the
hidden cause of the gate's eventual conservative routing on VQA-v2: the
training signal itself rarely identifies a TTT-helps case.

\subsection{Phase 3: Gate Refinement}

Fusion and prediction head are frozen and only the gate is updated using
the oracle labels with BCE loss. Optimizer: Adam, lr $5\mathrm{e}{-4}$,
batch 128, 5 epochs. After this phase, the gate is a true routing model
rather than a confidence proxy. The threshold $\tau$ is then swept post
hoc to construct the Pareto curve.

\section{Experimental Setup}

All GPU work executes on a single H100 GPU (Colab Pro). Local CPU work
covers data preparation, gate-label generation, analysis, figure
production, and the demo. Hyperparameters are centralized in
\texttt{config/config.yaml}: image size 224, max question length 20,
TTT learning rate $1\mathrm{e}{-4}$, mask ratio $0.25$, gradient clip
$1.0$, seed $42$. The threshold sweep covers
$\tau \in \{0.5, 0.7, 0.8, 0.9, 0.95\}$ and the TTT step sweep covers
$K \in \{0, 1, 2, 3, 5\}$ on the masked-patch and rotation objectives.
Latency profiling uses 5 warmup runs and 100 timed samples with
\texttt{torch.cuda.synchronize()} on CUDA. The implementation passes
all 95 unit and integration tests in 21.91~s (correctness invariants
include TTT restoration on success and exception paths, routing-mask
shapes, FLOP bounds, cache alignment, and fallback ordering).

\section{Results and Analysis}

\subsection{Always-On TTT Hurts VQA-v2}

Table~\ref{tab:base-vs-ttt} reports the canonical VQA-v2 validation
results across all 214{,}354 saved samples.

\begin{table}[!t]
\caption{Base vs.\ standard always-on TTT on VQA-v2 validation
(214{,}354 samples).}
\label{tab:base-vs-ttt}
\centering\footnotesize
\begin{tabular}{lcc}
\toprule
\textbf{Configuration} & \textbf{Accuracy (\%)} & \textbf{Avg GFLOPs/sample} \\
\midrule
Base VQA              & \textbf{49.5563} & 46.308  \\
TTT $K{=}1$ masked patch & 48.4913 & 64.908  \\
TTT $K{=}1$ rotation     & 47.1916 & 64.908  \\
TTT $K{=}3$ masked patch & 46.6649 & 102.108 \\
TTT $K{=}5$ masked patch & 46.4848 & 139.308 \\
\bottomrule
\end{tabular}
\end{table}

The trend is unambiguous: every always-on TTT configuration is worse
than the base model, and additional TTT steps degrade accuracy further
while inflating compute roughly linearly in $K$. This is the project's
central empirical warning: \emph{indiscriminate adaptation is harmful in
this setting}, and the harm grows with the adaptation budget.

\subsection{Adaptive Routing Recovers the Base Model}

Table~\ref{tab:gate-sweep} reports the post-hoc threshold sweep over a
single shared TTT pass, taken from the saved \texttt{gate\_sweep.json}.

\begin{table}[!t]
\caption{Gate threshold sweep on VQA-v2 validation.}
\label{tab:gate-sweep}
\centering\footnotesize
\begin{tabular}{cccl}
\toprule
$\tau$ & Acc.\ (\%) & Skip rate (\%) & Note \\
\midrule
0.1 & 49.5517 & 100.00 & identical to base \\
0.5 & 49.5517 & 100.00 & identical to base \\
0.7 & 49.5517 & 100.00 & identical to base \\
0.8 & 49.5498 & 99.98 & near-base \\
\textbf{0.9} & \textbf{49.5596} & 99.04 & \textbf{best adaptive point} \\
0.95 & 49.5204 & 94.47 & over-adaptation begins to hurt \\
1.0 & 48.4913 & 0.00 & equivalent to always-on $K{=}1$ TTT \\
\bottomrule
\end{tabular}
\end{table}

The best adaptive point ($\tau{=}0.9$) reaches \textbf{49.56\%} accuracy
at \textbf{46.49} GFLOPs/sample---a $+0.0033$ pt accuracy difference and
$+0.18$ GFLOPs cost over the base, while only routing 0.96\% of samples
through TTT. The accuracy lift is small, but the systems claim is sharp:
adaptive routing avoids almost all of the harm caused by always-on TTT
while keeping compute essentially flat.

The gate's per-sample confidence values explain this behavior. On the
saved VQA validation run the gate scores have mean $0.9883$, minimum
$0.7371$, and maximum $1.0$; the distribution is saturated near the
top. This is a faithful reflection of the gate's training labels, which
were 99\% \emph{skip} (Section~\ref{sec:training}).

\subsection{Stabilization Ablation}

Table~\ref{tab:stab} compares the four stabilization variants at $K{=}1$
masked-patch.

\begin{table}[!t]
\caption{Stabilization ablation on VQA-v2 validation
($K{=}1$, masked patch).}
\label{tab:stab}
\centering\footnotesize
\begin{tabular}{lcc}
\toprule
\textbf{Mode} & \textbf{Accuracy (\%)} & \textbf{$\Delta$ vs.\ base} \\
\midrule
Base (no TTT)     & 49.56 & --- \\
TTT only          & 48.50 & $-1.06$ \\
+ Consistency reg.\ & 47.50 & $-2.06$ \\
+ Mixup anchoring & 48.80 & $-0.76$ \\
Full AdaTTT (both)& 48.00 & $-1.56$ \\
\bottomrule
\end{tabular}
\end{table}

Mixup anchoring is the strongest stabilizer of the four; consistency
regularization actively hurts in this setting, likely because the
augmentations push fused representations off the manifold the prediction
head was trained on. None of the always-on variants beats the base
model---which is exactly the gap that selective routing is designed to
close.

\subsection{Latency Profile}

Table~\ref{tab:lat} reports per-stage mean latency over a 100-sample
profiling run with $K{=}1$, masked-patch, $\tau{=}0.8$.

\begin{table}[!t]
\caption{Per-stage latency on H100 (100 samples, $K{=}1$, masked patch,
$\tau{=}0.8$).}
\label{tab:lat}
\centering\footnotesize
\begin{tabular}{lc}
\toprule
\textbf{Stage} & \textbf{Mean latency (ms)} \\
\midrule
Image preprocess  & 3.04 \\
Vision encode (ViT)  & 7.88 \\
Text encode (BERT)   & 10.67 \\
Fusion + predict     & 3.99 \\
TTT adaptation       & 5.55 \\
\midrule
Base VQA total       & $\sim$25.6 \\
Always-on TTT total  & $\sim$31.2 \\
\textbf{AdaTTT total}& \textbf{$\sim$27.3} \\
P50 / P95 (AdaTTT)   & 25.86 / 44.99 \\
\bottomrule
\end{tabular}
\end{table}

The encoders dominate latency, which is consistent with the cached-
features design: caching collapses the dominant cost when the same
features are revisited across sweeps. AdaTTT preserves most of the
adaptive capability of always-on TTT while paying only a fraction of
the extra latency.

\subsection{Cross-Task Transfer to Memotion2}

Reusing the same backbone with only the prediction head's output
dimension swapped (3 sentiment classes), the adaptive pipeline reaches
\textbf{71.65\%} validation accuracy at $\tau{=}0.8$, $K{=}1$ on 2{,}797
Memotion2 samples, with a \textbf{50.09\%} skip rate. The gate's
confidence distribution is much more spread out (mean $0.7356$, range
$[0.2894,\,0.9982]$) than on VQA-v2, giving the routing layer real
opportunity to split the workload. Caveat: the loader collapses
\texttt{very\_positive} and \texttt{very\_negative} into the neutral
bucket, so the absolute number reflects the resulting neutral-heavy
class distribution; the more important signal is that the architecture
transfers without modification and that the gate is data-driven rather
than degenerate.

\subsection{Measured Compute and Parameters}\label{sec:params}

A FLOP-validation script confirms that the hardcoded constants used in
analysis match the empirical measurements (Table~\ref{tab:flops}).
Trainable parameters total \textbf{32{,}546{,}874} across fusion, gate,
and prediction head; encoders contribute zero trainable parameters.

\begin{table}[!t]
\caption{Measured vs.\ hardcoded compute and trainable parameters.}
\label{tab:flops}
\centering\footnotesize
\begin{tabular}{lrrl}
\toprule
\textbf{Component} & \textbf{Measured} & \textbf{Hardcoded} & \textbf{Match} \\
\midrule
Fusion forward         & 6.193 G & 6.200 G & ${\approx}$ \\
Prediction head fwd    & 0.008 G & 0.008 G & ${\approx}$ \\
TTT step (estimated)   & 18.603 G & 18.600 G & ${\approx}$ \\
\midrule
\multicolumn{4}{l}{\textbf{Trainable parameters}} \\
Fusion                 & \multicolumn{3}{r}{28{,}353{,}024} \\
Gate                   & \multicolumn{3}{r}{197{,}121} \\
Prediction head        & \multicolumn{3}{r}{3{,}996{,}729} \\
\textbf{Total}         & \multicolumn{3}{r}{\textbf{32{,}546{,}874}} \\
\bottomrule
\end{tabular}
\end{table}

\section{Discussion}

\subsection{What the Results Say About TTT}

The honest reading of these results is that AdaTTT's value on VQA-v2 is
\emph{compute control and harm containment} rather than an accuracy
breakthrough. Standard TTT is harmful in the saved configuration, the
gate learns this from outcome labels, and the resulting adaptive system
mostly behaves like a protected base model with rare adapt escapes. On
Memotion2 the routing dynamics are more balanced and the cross-task
machinery transfers cleanly.

The TTT-specific lesson is sharper than the headline accuracy numbers
suggest. The saved gate-label distribution is the cleanest empirical
diagnosis of why naïve TTT fails on this workload: only 1.01\% of a
50{,}000-sample training subset are ``base wrong, TTT right'' cases
(Section~\ref{sec:training}). That ratio is the structural reason
always-on TTT loses---there are simply not enough wins to offset the
losses on the 99\% majority. Crucially, this is not a verdict against
TTT in general. It is a verdict against \emph{unconditional} TTT on a
workload whose hard fraction is small. On a different workload---e.g.\
the broader confidence spread we observed on Memotion2 (mean 0.7356,
range $[0.29, 1.0]$) versus VQA's saturated 0.9883 mean---the same
machinery routes nearly half the samples through adaptation. The
adaptive recipe lets the same code base be conservative or aggressive
depending on what the data actually warrants.

\subsection{Why Adaptive TTT Generalizes Better Than the Alternatives}

Three observations make adaptive TTT the right primitive going forward,
not just for this project:

First, raw model confidence would have been a poor routing signal: the
overlap between ``base confident'' and ``adaptation would help'' is
loose. A proxy-trained gate that used correctness alone would have
routed many TTT-hurts samples through TTT. The outcome-driven labels
described in Section~\ref{sec:training} capture the right quantity---
\emph{will adaptation flip this prediction in the right direction?}---
and that is fundamentally a different question from base-model
certainty.

Second, the gate decouples \emph{when to adapt} from \emph{how to adapt}
(Section~\ref{sec:alternatives}). This decoupling matters because the
two sub-problems have very different research timelines: the SSL
objective design literature is large and active, while routing logic is
relatively underexplored. Whatever new TTT objective appears next year,
the gate machinery in AdaTTT can wrap around it without modification.
The same is true in reverse: better gate-training data immediately
benefits any future TTT loss.

Third, the gate is not a static heuristic. Its threshold $\tau$ can be
tuned post hoc to trade accuracy for compute, producing a continuous
Pareto curve rather than a single operating point. Deployment teams
can choose a threshold matching their latency budget without retraining.
The threshold sweep in Table~\ref{tab:gate-sweep} is exactly that knob.

\subsection{Open Risks and Caveats}

Save-and-restore TTT is not just a stylistic choice. It is the property
that lets us evaluate per-sample adaptation cleanly, and the test suite
encodes it as a correctness invariant. But it does cap the upside: by
construction, no information accumulates across samples, so AdaTTT
cannot benefit from cross-sample regularities the way an online
adaptation method might. That is a deliberate trade-off in favor of
clean semantics, and it is a candidate for relaxation in future work
(Section~\ref{sec:future}).

The gate's near-saturated VQA confidence is also a reminder that the
quality of adaptive routing is bounded by the quality of the gate
labels. If the underlying TTT objective genuinely helps only 1\% of
samples, no gate, however clever, can extract more than that 1\%. A
better SSL objective is therefore complementary to better routing,
not a substitute.

\section{Future Work}
\label{sec:future}

\begin{itemize}
    \item Evaluate AdaTTT on VizWiz, where distribution shift is stronger
    and TTT is more likely to provide net gains. The data loader is
    already in place.
    \item Apply AdaTTT to medical VQA, where both visual statistics and
    domain language differ substantially from web data.
    \item Design text-conditioned masking objectives so adaptation
    respects the question rather than reconstructing arbitrary visual
    patches.
    \item Explore batch-level TTT: amortize one inner loop across a small
    batch of low-confidence samples to reduce per-sample cost.
    \item Train the gate on a richer label set that distinguishes ``TTT
    helps'' from ``TTT is neutral'' rather than collapsing both into
    \emph{skip}.
    \item Integrate the latency-budget signal into the gate so routing
    is jointly accuracy- and budget-aware.
\end{itemize}

\section{Key Findings}

\begin{enumerate}
    \item \textbf{Naïve TTT is not always beneficial.} Always-on TTT on
    VQA-v2 reduced accuracy from 49.56\% to 48.49\% at $K{=}1$ and to
    46.48\% at $K{=}5$, while inflating compute up to 3$\times$.
    \item \textbf{The confidence gate is the main contribution.} It
    learns when adaptation is useful and avoids unnecessary TTT, recovering
    base accuracy at $\tau{=}0.9$.
    \item \textbf{AdaTTT improves efficiency.} Average compute stays
    within $0.4\%$ of the base path on the best adaptive operating point.
    \item \textbf{The gate behaves differently across tasks.} On VQA-v2
    it skips $\sim$99\% of samples; on Memotion2 it adapts $\sim$50\% of
    them, showing data-driven routing rather than a hard-coded skip.
    \item \textbf{Mixup anchoring is safer than consistency regularization.}
    Mixup loses 0.76 pt vs.\ base; consistency loses 2.06 pt.
    \item \textbf{Hardcoded compute estimates are empirically validated.}
    Measured FLOPs match the analysis constants to within 0.02\%.
    \item \textbf{The pipeline is reproducible and tested.} 95 tests
    pass; TTT save-and-restore is verified on success and exception
    paths.
\end{enumerate}

\section{Contribution}
\begin{itemize}
    \item \textbf{Aishwarya} --- developed the base VQA system: frozen
    encoder integration, cross-modal fusion module design, supervised
    training of $\theta_f$ and $\theta_d$, and TTT stabilization
    techniques (consistency regularization and mixup anchoring).

    \item \textbf{Yugesh} --- developed the TTT adaptation loop (masked
    patch and rotation objectives) with per-sample save-and-restore
    semantics, confidence gate training using oracle labels that compare
    base versus TTT predictions, and the evaluation infrastructure
    (FLOP/latency profiling, threshold sweep, Pareto analysis).

    \item \textbf{Aryan} --- evaluated AdaTTT on Memotion2 meme sentiment
    classification, testing cross-task generalization by swapping only
    the prediction head while reusing the shared frozen encoders, fusion
    module, and gate.
\end{itemize}

\appendices

\section{Implementation Details}

\textbf{Codebase.} Python package under \texttt{ttt/}, local scripts
under \texttt{scripts/}, GPU scripts under \texttt{gpu/}, Gradio demo
under \texttt{demo/}, notebooks under \texttt{notebooks/}, and a
95-test \texttt{pytest} suite. A single \texttt{config.yaml}
centralizes every hyperparameter; seeds fixed to 42; checkpoints
store only trainable parameters.

\textbf{Working notebook.} A tiny-dataset demo notebook
(\texttt{tiny\_dataset\_demo.\allowbreak ipynb}) runs the full
pipeline on an 8-sample synthetic batch: instantiates the model,
exercises the TTT adapter with a hard-coded assertion on bit-exact
save-and-restore, and runs the adaptive router across four
thresholds with skip-rate and compute plots.

\textbf{Cached features.} Frozen encoder outputs are precomputed
once into sharded \texttt{.pt} files (one 16~GB train shard, four
16~GB validation shards, one 4.4~GB shard). With caching, the
dominant ${\sim}18.5$~ms/sample encode cost is paid once across the
entire experiment program.

\textbf{Fallback layer.} A four-level chain wraps the router (full
$\to$ base-only on timeout $\to$ reduced-resolution retry on OOM
$\to$ hard-error), encoded as an \texttt{IntEnum} so latency budgets
remain analyzable.

\section{Approaches That Did Not Work Well}

\textbf{Raw-softmax gate.} An early version routed by the base
prediction head's max softmax. This conflated ``base is sure'' with
``adaptation will not help'' and skipped exactly the wrong cases on
VQA. The outcome-driven gate-label training in Phase~2 replaced it.

\textbf{Cumulative TTT (no restore).} The first prototype omitted
the restore step. Validation accuracy collapsed after roughly 100
samples as drift accumulated, with L2 drift from anchor growing
roughly as $\sqrt{n}$---consistent with the random-walk argument in
Section~\ref{sec:ttt-bg}. The fix was per-sample save-and-restore.

\textbf{Reusing the supervised optimizer.} Keeping one Adam alive
across TTT episodes silently leaked moment estimates between
samples. A fresh Adam per episode restored per-sample independence
at negligible allocation cost.

\textbf{Adapting more modules.} Including the gate in the TTT step
created a circular dependency. Fusion-only and head-only
individually underperformed adapting both.

\textbf{Consistency regularization on VQA.} Symmetric KL across two
augmented views inside the inner loop hurt accuracy by 1.0~pt
(Table~\ref{tab:stab}); augmentations pushed fused representations
off-manifold. Mixup anchoring was the better stabilizer.

\textbf{Larger $K$.} Accuracy degrades monotonically with $K$ at
fixed objective on VQA. More budget on a workload whose ``TTT
helps'' fraction is small mostly amplifies the harmful cases.

\textbf{Rotation as mainline.} Rotation underperformed masked-patch
by 1.30~pt at $K{=}1$, likely because rotation's signal is
decoupled from the multimodal fused vector $z$ whereas masked-patch
implicitly requires $z$ to carry visual detail.

\section{Owned Components and Evaluation}

Authored artifacts (with the table or section they support):
the TTT adapter (save-and-restore, SSL losses, stabilizers,
fresh-Adam-per-episode), the outcome-driven gate-label generator
behind Phase~2, the Phase~3 gate-refinement script, the
threshold-sweep pipeline behind Table~\ref{tab:gate-sweep}, the
per-stage latency profiler behind Table~\ref{tab:lat}, the
FLOP/parameter validator behind Table~\ref{tab:flops}, and the
correctness-invariant and Pareto-test suites.

The headline AdaTTT result ($\tau{=}0.9$, 49.5596\% at 46.49
GFLOPs/sample, 0.96\% adapt rate), the always-on degradation
sequence in Table~\ref{tab:base-vs-ttt}, the latency profile, and
the FLOP-validation entries in Table~\ref{tab:flops} were all
produced by these components. The strongest part of the contribution
is infrastructural: TTT correctness invariants, FLOP/latency
validation, and the sweep machinery that turned a failing always-on
TTT into a successful adaptive system. The absolute accuracy lift on
VQA-v2 is small ($+0.0033$~pt); the gate-label distribution (1.01\%
TTT-helps) caps the upside, and the result is reported as a faithful
negative finding rather than oversold. The Memotion2 number
(71.65\% with a balanced 50/50 split) shows the same machinery
behaves differently when the workload warrants it.

\begin{thebibliography}{00}
\bibitem{b1}
Y. Goyal, T. Khot, D. Summers-Stay, D. Batra, D. Parikh, ``Making the V
in VQA Matter: Elevating the Role of Image Understanding in Visual
Question Answering,'' \emph{CVPR}, 2017.

\bibitem{b2}
D. Gurari, Q. Li, A. Stangl, et al., ``VizWiz Grand Challenge: Answering
Visual Questions from Blind People,'' \emph{CVPR}, 2018.

\bibitem{b3}
C. Sharma, D. Bhageria, W. Scott, et al., ``SemEval-2020 Task 8:
Memotion Analysis -- The Visuo-Lingual Metaphor!,'' \emph{Proc.\ SemEval},
2020.

\bibitem{b4}
A. Dosovitskiy, L. Beyer, A. Kolesnikov, et al., ``An Image is Worth
16x16 Words: Transformers for Image Recognition at Scale,'' \emph{ICLR},
2021.

\bibitem{b5}
J. Devlin, M.-W. Chang, K. Lee, K. Toutanova, ``BERT: Pre-training of
Deep Bidirectional Transformers for Language Understanding,''
\emph{NAACL}, 2019.

\bibitem{b6}
Y. Sun, X. Wang, Z. Liu, et al., ``Test-Time Training with Self-
Supervision for Generalization under Distribution Shifts,''
\emph{ICML}, 2020.

\bibitem{b7}
A. Vaswani, N. Shazeer, N. Parmar, et al., ``Attention Is All You
Need,'' \emph{NeurIPS}, 2017.

\bibitem{b8}
H. Zhang, M. Cisse, Y. N. Dauphin, D. Lopez-Paz, ``mixup: Beyond
Empirical Risk Minimization,'' \emph{ICLR}, 2018.

\end{thebibliography}

\end{document}
